The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2017-01-21 script/en/chapters/bitciphers.tex
CT Book ed 12: Small text correction in bitciphers.tex
CT Book ed 12: Deleted some comments; added Python to Index.

Collections of bits are denoted by several different names\footnote{%
   that all refer the same conceptual entities. However in Python or SageMath
   the different names sometimes denote different types.
},
depending on the context:
vectors\index{vector}, lists\index{list}, ($n$-) tuples\index{tuple},
\ldots\ For certain sizes we often use special denotations such as
bytes\index{byte} or octets\index{octet} (for 8 bits), words\index{word}
(for 32 or 64 bytes depending on the processor architecture) \ldots\ 
In the present chapter we usually use the denomination ``bitblocks\index{bitblock}''
that is common in cryptography.
Thus a {\bf bitblock} of length $n$ is a list $(x_1, \ldots, x_n)$ of
bits where the order matters. There are eight different bitblocks of length $3$:
\[
   (0,0,0), (0,0,1), (0,1,0), (0,1,1), (1,0,0), (1,0,1), (1,1,0), (1,1,1).
\]
Sometimes, if the danger of misunderstanding is negligeable, we write them as
bitstrings\index{bitstring}\footnote{%
   sometimes also as columns, that is $n \times 1$ matrices when the
   focus is on the interpretation as vectors\index{vector}
}
without parentheses or commas:
\begin{equation}\label{eq-lexi}
   000, 001, 010, 011, 100, 101, 110, 111.
\end{equation}
We often use the abbreviation $x$ for $(x_1, \ldots, x_n)$. This short form 
highlights the fact that we consider bitblocks\index{bitblock} as objects
of their own.

@@

We used the term bitblock\index{bitblock} for a variety of slightly different
objects. A bitblock $b = (b_1, \ldots, b_n) \in \F_2^n$ describes:
\begin{itemize}
   \item a vector\index{vector} $b \in \F_2^n$ written as a row or a column.
      This is the primary meaning of the term bitblock\footnote{%
         in Python/SageMath implemented as list\index{list}, or tuple, or vector
      }.
   \item an argument of a Boolean function or map of $n$ variables, also used
      as row index of a value table (or truth table\index{truth table})
   \item a bitstring\index{bitstring} of length $n$
   \item a subset $I \subseteq \{1, \ldots, n\}$ defined by $b$
      as indicator: $i \in I \Leftrightarrow b_i = 1$
   \item a linear form\index{linear form} $\alpha$ on $\F_2^n$ expressed as
      sum of the variables $x_i$ with $b_i = 1$. The evaluation of $\alpha$
      comes down to the scalar product of vectors: $\alpha(x) = b \cdot x$.
   \item a monomial\index{monomial} in $n$ variables $x_1, \ldots, x_n$ with all
      partial degrees $\leq 1$. In this interpretation $b_i$ specifies the
      exponent $0$ or $1$ of the variable $x_i$.
   \item an integer between $0$ and $2^n - 1$ in binary representation (that
      is in the base-2 system). The sequence of binary ``digits'' (= bits)
      is identical with the corresponding bitstring\footnote{%
         The SageMath method {\tt binary()} transforms an integer to a bitstring,
         suppressing leading zeros. Example: {\tt 10.binary()} yields {\tt '1010'}.
      }. Conversely the integer is the index (beginning with 0) of the bitstring
      when the biststrings are lexicographically ordered in a list.
\end{itemize}
Of course there are further interpretations---after all each piece of
information has a binary coding. The bitblocks for $n = 3$ are listed
in Table~\ref{tab-bool-3}. SageMath sample~\ref{ss-bool-conv} has some transformation
functions.

@@

In this way, the bitblock $(0,0,0,0,0,1,1,1)$ or, even more parsimoniously,
the bitstring
\[
     00000111
\]
of length $2^3 = 8$ completely specifies the truth table of $f_0$.\footnote{%
  In Python/SageMath bitblocks are implemented as lists\index{list}.
}

@@

For an implementation of Boolean functions in SageMath (or Python) see
Appendix~\ref{ss-bool-class} (class {\tt BoolF()}). SageMath itself has a class
{\tt sage.crypto.boolean\_function} that contains many of the needed
methods, also the conversion of a truth table\index{truth table} into ANF.
Here we describe an independent implementation.

@@

As a slightly more complex
example we define a toy cipher ``Mini-Lucifer\index{Mini-Lucifer}''
that employs the S-boxes and a permutation of the true
{\sc Lucifer}\index{Lucifer}. Here is the construction, see
Figure~\ref{fig-bool-miniLuc}:
\begin{itemize}
   \item Before and after each round map we add a partial key. We
      use two keys $k^{(0)}$ and $k^{(1)}$ in alternating order.
      They consist of the first or last 8 bits of the 16 bit master
      key. In particular for $r \geq 3$ the round keys\index{round key}
      are not independent.
   \item The round function consists of a parallel arrangement of the
      two S-boxes, as in the example of Section~\ref{ss-bool-par},
      followed by the permutation $\textrm{P}$.
   \item The permutation $\textrm{P}$ maps a single byte (octet) to
      itself as defined in SageMath sample~\ref{Sage-code-LucP}. As usual
      for SP-networks we omit it in the last round.
\end{itemize}
SageMath sample~\ref{Sage-code-miniLuc} contains the SageMath/Python code.

@@

For our 2-round example with Mini-Lucifer\index{Mini-Lucifer} the
systematic search is feasible by pencil and paper or by a SageMath or
Python script. Our example has the following characteristics:
\begin{itemize}
   \item $\alpha = (\alpha_1,\alpha_2)$\footnote{%
      $\alpha_1$ was formerly denoted $\alpha$. Now for uniformity we
      make both components of all linear forms explicit and index them
      by $1$ and $2$.
      } with $\alpha_1 \:\hat{=}\: 1 \:\hat{=}$ \verb:0001: and
      $\alpha_2 \:\hat{=}\: 0 \:\hat{=}$ \verb:0000:
   \item $\beta = (\beta_1,\beta_2)$ with $\beta_1 \:\hat{=}\: 13 \:\hat{=}$
      \verb:1101: and $\beta_2 \:\hat{=}\: 0 \:\hat{=}$ \verb:0000:
   \item $\beta' = (\beta_1',\beta_2')$ with $\beta_1' \:\hat{=}\: 1\:\hat{=}$
      \verb:0001:, $\beta_2' \:\hat{=}\: 12 \:\hat{=}$ {\tt 1100}
   \item $\gamma = (\gamma_1,\gamma_2)$ with $\gamma_1 \:\hat{=}\: 13 \:\hat{=}$
       \verb:1101:, $\gamma_2 \:\hat{=}\: 6 \:\hat{=}$ \verb:0110:
   \item $\tau_1 = \frac{3}{4}$, $\tau_2' = \frac{3}{4}$,
      $\tau_2'' = \frac{1}{2}$, $\tau_2 = \frac{3}{8}$,
      $\tau = \frac{9}{32}$, $p = \frac{41}{64} = 0,640625$
   \item $c_0 + c_1 + c_3 + c_5 + c_6 \stackrel{p}{\approx}
      a_3 + k_0 + k_1 + k_5 + k_6 + k_{11} + k_{12} + k_{13}$
\end{itemize}
An alternative choice of $\gamma_2$ is $\gamma_2 \:\hat{=}\: 14 \:\hat{=}$
\verb:1110:; this yields a linear path with the characteristics
\begin{itemize}
   \item $\alpha \:\hat{=}\: (1,0)$, $\beta \:\hat{=}\: (13,0)$, $\beta' \:\hat{=}\: (1,12)$,
      $\gamma \:\hat{=}\: (13,14)$
   \begin{itemize}
      \item $\tau = -\frac{9}{32}$, $p = \frac{23}{64} = 0,359375$
      \item $c_0 + c_1 + c_3 + c_4 + c_5 + c_6 \stackrel{p}{\approx}
         a_3 + k_0 + k_1 + k_4 + k_5 + k_6 + k_{11} + k_{12} + k_{13}$
   \end{itemize}
\end{itemize}
The systematic search finds two even ``better'' linear paths,
characterized by
\begin{itemize}
   \item $\alpha \:\hat{=}\: (8,0)$, $\beta \:\hat{=}\: (8,0)$, $\beta' \:\hat{=}\: (1,0)$,
      $\gamma \:\hat{=}\: (13,0)$
   \begin{itemize}
      \item $\tau = -\frac{3}{8}$, $p = \frac{5}{16} = 0,3125$
      \item $c_0 + c_1 + c_3 \stackrel{p}{\approx} a_0 + k_1 + k_3 + k_{11}$
   \end{itemize}
   \item $\alpha \:\hat{=}\: (15,0)$, $\beta \:\hat{=}\: (8,0)$, $\beta' \:\hat{=}\: (1,0)$,
      $\gamma \:\hat{=}\: (13,0)$
   \begin{itemize}
      \item $\tau = -\frac{3}{8}$, $p = \frac{5}{16} = 0,3125$
      \item $c_0 + c_1 + c_3 \stackrel{p}{\approx} a_0 + a_1 + a_2 + a_3 + k_2 + k_{11}$
   \end{itemize}
\end{itemize}
that do not completely exhaust the potential of the single S-boxes but
on the other hand activate only one S-box of the second round, and thereby
show the larger potential $\lambda = \frac{9}{64}$. Thus we get a 95\%
success probability with only
\[
     N = \frac{3}{\lambda} = \frac{64}{3} \approx 21
\]
known\index{known plaintext}\index{plaintext!known} plaintexts for determining
one bit.

@@

sage: testtext = "Come at nine"
sage: bintext = txt2bbl(testtext)
sage: binstr = bbl2str(bintext); binstr
'010000110110111101101101011001010010000001100001
011101000010000001101110011010010110111001100101'
sage: testkey = [1,0,0,1,0,1,1,0]
sage: keystr = bbl2str(testkey); keystr
'10010110'
sage: cipher = xor(bintext,testkey)
sage: ciphstr = bbl2str(cipher); ciphstr
'110101011111100111111011111100111011011011110111
111000101011011011111000111111111111100011110011'
\end{verbatim}
\caption{XOR encryption\index{XOR}\index{encryption!XOR}\index{cipher!XOR} in Python/SageMath}\label{Sage-code-bool-xor}
\end{sagecode}

@@

def fsr(f,x,n):
  """Generate a feedback shift register sequence.
  Parameters: Boolean function f, start vector x,
  number n of output bits."""
  u = x
  outlist = []
  for i in range (0,n):
    b = f.valueAt(u)
    c = u.pop()
    u.insert(0,b)
    outlist.append(c)
  return outlist
\end{verbatim}
\caption{A feedback shift register (FSR) in Python/SageMath}\label{Sage-code-bool-fsr}
\end{sagecode}

@@

sage: bits = "1000010111001001"
sage: x = str2bbl(bits)
sage: f = BoolF(x)
sage: start = [0,1,1,1]
sage: bitlist = fsr(f, start, 32)
sage: print(bbl2str(bitlist))
11101010101010101010101010101010
\end{verbatim}
\caption{A (very poor) pseudo-random sequence\index{pseudo-random sequence}
   in Python/SageMath}\label{Sage-code-bool-fsr1}
\end{sagecode}

@@

def lfsr(s,x,n):
  """Generate a linear feedback shift register sequence.
  Parameters: Coefficient vector s, start vector x, number n of
  output bits."""
  l = len(s)
  assert l == len(x), "lfsr_Error: Bad length of start vector."
  u = x                           # in Python use u = x.copy()
  outlist = []
  for i in range (0,n):
    b = binScPr(s, u)
    c = u.pop()
    u.insert(0,b)
    outlist.append(c)
  return outlist
\end{verbatim}
\caption{Defining an LFSR\index{LFSR} in Python/SageMath}\label{Sage-code-bool-lfsr}
\end{sagecode}

@@

sage: coeff = [0,1,1,0,1,0,0,0,0,0,0,0,0,0,0,1]
sage: start = [0,1,1,0,1,0,1,1,0,0,0,1,0,0,1,1]
sage: bitlist = lfsr(coeff, start, 1024)
sage: print(bitlist)
### Visualization
sage: m = matrix(GF(2),32,bitlist)
sage: row = [1]*32
sage: n = matrix(GF(2),32,row*32)  # All entries 1
sage: p = m+n                  # Toggle bits of m ---> 1 = black
sage: p.subdivide(range(0,33),range(0,33))
sage: matrix_plot(p, subdivisions=True)
\end{verbatim}
\caption{A pseudo-random bit sequence\index{pseudo-random sequence}
   in Python/SageMath}\label{Sage-code-bool-lfsr1}
\end{sagecode}

@@

\begin{description}
   \item[Class {\tt LSFR}:] ~
      \begin{description}
         \item[Attributes:] ~
            \begin{itemize}
               \item {\tt length}: the length of the register
               \item {\tt taplist} (constant): the list of coefficients (or taps)
                  that define the bits for feedback
               \item {\tt state} (variable): the state of the register
            \end{itemize}
         \item[Methods:] ~
            \begin{itemize}
               \item {\tt setLength}: define the length (used only implicitly
                  for initialization)
               \item {\tt setTaps}: define the list of taps (used only implicitly
                  for initialization)
               \item {\tt setState}: set the state of the register
               \item {\tt getLength}: output the length
               \item {\tt nextBits}: generate a given number of output bits,
                  and set the next state
            \end{itemize}
      \end{description}
\end{description}
For observing the register a method (generically called {\tt \_\_str\_\_}
in Python) is convenient that outputs the attributes in human-readable form.

@@

But implementing a useful bitstream cipher takes more than just a good random generator:
\begin{itemize}
   \item Message integrity\index{integrity} requires additional means
      such as a combination with a cryptographic hash function\index{hash function}. 
   \item The operational conditions must prevent the reuse of (parts of) the
      key stream\index{key stream} in a reliable way. This means that the
      key management\index{key management} requires utmost prudence.
      A possible approach\footnote{%
         This was a usual approach with the cipher machines of World War II.
      }
      is using a longtime general key that consists of certain inner parameters
      of the random generator, and use the remaining parameters including the
      initial state as one-time message key\index{OTP}.
\end{itemize}

@@

SageMath has several functions for bitblocks\index{bitblock} and Boolean
functions\index{Boolean function}\index{function!Boolean}. These are
scattered over different modules, use different data types, and have
some limitations. We could live with them. Nevertheless in this text
we mostly use our own independent (and free) implementation\footnote{%
   This is known as ``reinventing the wheel''. Forgive it in view of
   intended uniformity and consistency. Working through the functions in the
   SageMath samples should facilitate learning the algorithms. A further aspect:
   Our modules are written in pure Python and may be used without installing
   SageMath.
}.

@@

The SageMath classes and functions defined in the following can be found in the
module \verb:bitciphers.sage:. They were developed with Python 3.4.1\footnote{%
   The only relevant difference with Python 2 is the operator {\tt //} for the
   division of integers.
} under OS X, and tested with SageMath 6.2.

>>>>

Collections of bits are denoted by several different names\footnote{%
   that all refer the same conceptual entities. However in Python\index{Python} or SageMath
   the different names sometimes denote different types.
},
depending on the context:
vectors\index{vector}, lists\index{list}, ($n$-) tuples\index{tuple},
\ldots\ For certain sizes we often use special denotations such as
bytes\index{byte} or octets\index{octet} (for 8 bits), words\index{word}
(for 32 or 64 bytes depending on the processor architecture) \ldots\ 
In the present chapter we usually use the denomination ``bitblocks\index{bitblock}''
that is common in cryptography.
Thus a {\bf bitblock} of length $n$ is a list $(x_1, \ldots, x_n)$ of
bits where the order matters. There are eight different bitblocks of length $3$:
\[
   (0,0,0), (0,0,1), (0,1,0), (0,1,1), (1,0,0), (1,0,1), (1,1,0), (1,1,1).
\]
Sometimes, if the danger of misunderstanding is negligeable, we write them as
bitstrings\index{bitstring}\footnote{%
   sometimes also as columns, that is $n \times 1$ matrices when the
   focus is on the interpretation as vectors\index{vector}
}
without parentheses or commas:
\begin{equation}\label{eq-lexi}
   000, 001, 010, 011, 100, 101, 110, 111.
\end{equation}
We often use the abbreviation $x$ for $(x_1, \ldots, x_n)$. This short form 
highlights the fact that we consider bitblocks\index{bitblock} as objects
of their own.

@@

We used the term bitblock\index{bitblock} for a variety of slightly different
objects. A bitblock $b = (b_1, \ldots, b_n) \in \F_2^n$ describes:
\begin{itemize}
   \item a vector\index{vector} $b \in \F_2^n$ written as a row or a column.
      This is the primary meaning of the term bitblock\footnote{%
         in Python/SageMath\index{Python} implemented as list\index{list}, or tuple, or vector
      }.
   \item an argument of a Boolean function or map of $n$ variables, also used
      as row index of a value table (or truth table\index{truth table})
   \item a bitstring\index{bitstring} of length $n$
   \item a subset $I \subseteq \{1, \ldots, n\}$ defined by $b$
      as indicator: $i \in I \Leftrightarrow b_i = 1$
   \item a linear form\index{linear form} $\alpha$ on $\F_2^n$ expressed as
      sum of the variables $x_i$ with $b_i = 1$. The evaluation of $\alpha$
      comes down to the scalar product of vectors: $\alpha(x) = b \cdot x$.
   \item a monomial\index{monomial} in $n$ variables $x_1, \ldots, x_n$ with all
      partial degrees $\leq 1$. In this interpretation $b_i$ specifies the
      exponent $0$ or $1$ of the variable $x_i$.
   \item an integer between $0$ and $2^n - 1$ in binary representation (that
      is in the base-2 system). The sequence of binary ``digits'' (= bits)
      is identical with the corresponding bitstring\footnote{%
         The SageMath method {\tt binary()} transforms an integer to a bitstring,
         suppressing leading zeros. Example: {\tt 10.binary()} yields {\tt '1010'}.
      }. Conversely the integer is the index (beginning with 0) of the bitstring
      when the biststrings are lexicographically ordered in a list.
\end{itemize}
Of course there are further interpretations---after all each piece of
information has a binary coding. The bitblocks for $n = 3$ are listed
in Table~\ref{tab-bool-3}. SageMath sample~\ref{ss-bool-conv} has some transformation
functions.

@@

In this way, the bitblock $(0,0,0,0,0,1,1,1)$ or, even more parsimoniously,
the bitstring
\[
     00000111
\]
of length $2^3 = 8$ completely specifies the truth table of $f_0$.\footnote{%
  In Python/SageMath\index{Python} bitblocks are implemented as lists\index{list}.
}

@@

For an implementation of Boolean functions in SageMath (or Python\index{Python}) see
Appendix~\ref{ss-bool-class} (class {\tt BoolF()}). SageMath itself has a class
{\tt sage.crypto.boolean\_function} that contains many of the needed
methods, also the conversion of a truth table\index{truth table} into ANF.
Here we describe an independent implementation.

@@

As a slightly more complex
example we define a toy cipher ``Mini-Lucifer\index{Mini-Lucifer}''
that employs the S-boxes and a permutation of the true
{\sc Lucifer}\index{Lucifer}. Here is the construction, see
Figure~\ref{fig-bool-miniLuc}:
\begin{itemize}
   \item Before and after each round map we add a partial key. We
      use two keys $k^{(0)}$ and $k^{(1)}$ in alternating order.
      They consist of the first or last 8 bits of the 16 bit master
      key. In particular for $r \geq 3$ the round keys\index{round key}
      are not independent.
   \item The round function consists of a parallel arrangement of the
      two S-boxes, as in the example of Section~\ref{ss-bool-par},
      followed by the permutation $\textrm{P}$.
   \item The permutation $\textrm{P}$ maps a single byte (octet) to
      itself as defined in SageMath sample~\ref{Sage-code-LucP}. As usual
      for SP-networks we omit it in the last round.
\end{itemize}
SageMath sample~\ref{Sage-code-miniLuc} contains the SageMath/Python\index{Python} code.

@@

For our 2-round example with Mini-Lucifer\index{Mini-Lucifer} the
systematic search is feasible by pencil and paper or by a SageMath or
Python\index{Python} script. Our example has the following characteristics:
\begin{itemize}
   \item $\alpha = (\alpha_1,\alpha_2)$\footnote{%
      $\alpha_1$ was formerly denoted $\alpha$. Now for uniformity we
      make both components of all linear forms explicit and index them
      by $1$ and $2$.
      } with $\alpha_1 \:\hat{=}\: 1 \:\hat{=}$ \verb:0001: and
      $\alpha_2 \:\hat{=}\: 0 \:\hat{=}$ \verb:0000:
   \item $\beta = (\beta_1,\beta_2)$ with $\beta_1 \:\hat{=}\: 13 \:\hat{=}$
      \verb:1101: and $\beta_2 \:\hat{=}\: 0 \:\hat{=}$ \verb:0000:
   \item $\beta' = (\beta_1',\beta_2')$ with $\beta_1' \:\hat{=}\: 1\:\hat{=}$
      \verb:0001:, $\beta_2' \:\hat{=}\: 12 \:\hat{=}$ {\tt 1100}
   \item $\gamma = (\gamma_1,\gamma_2)$ with $\gamma_1 \:\hat{=}\: 13 \:\hat{=}$
       \verb:1101:, $\gamma_2 \:\hat{=}\: 6 \:\hat{=}$ \verb:0110:
   \item $\tau_1 = \frac{3}{4}$, $\tau_2' = \frac{3}{4}$,
      $\tau_2'' = \frac{1}{2}$, $\tau_2 = \frac{3}{8}$,
      $\tau = \frac{9}{32}$, $p = \frac{41}{64} = 0,640625$
   \item $c_0 + c_1 + c_3 + c_5 + c_6 \stackrel{p}{\approx}
      a_3 + k_0 + k_1 + k_5 + k_6 + k_{11} + k_{12} + k_{13}$
\end{itemize}
An alternative choice of $\gamma_2$ is $\gamma_2 \:\hat{=}\: 14 \:\hat{=}$
\verb:1110:; this yields a linear path with the characteristics
\begin{itemize}
   \item $\alpha \:\hat{=}\: (1,0)$, $\beta \:\hat{=}\: (13,0)$, $\beta' \:\hat{=}\: (1,12)$,
      $\gamma \:\hat{=}\: (13,14)$
   \begin{itemize}
      \item $\tau = -\frac{9}{32}$, $p = \frac{23}{64} = 0,359375$
      \item $c_0 + c_1 + c_3 + c_4 + c_5 + c_6 \stackrel{p}{\approx}
         a_3 + k_0 + k_1 + k_4 + k_5 + k_6 + k_{11} + k_{12} + k_{13}$
   \end{itemize}
\end{itemize}
The systematic search finds two even ``better'' linear paths,
characterized by
\begin{itemize}
   \item $\alpha \:\hat{=}\: (8,0)$, $\beta \:\hat{=}\: (8,0)$, $\beta' \:\hat{=}\: (1,0)$,
      $\gamma \:\hat{=}\: (13,0)$
   \begin{itemize}
      \item $\tau = -\frac{3}{8}$, $p = \frac{5}{16} = 0,3125$
      \item $c_0 + c_1 + c_3 \stackrel{p}{\approx} a_0 + k_1 + k_3 + k_{11}$
   \end{itemize}
   \item $\alpha \:\hat{=}\: (15,0)$, $\beta \:\hat{=}\: (8,0)$, $\beta' \:\hat{=}\: (1,0)$,
      $\gamma \:\hat{=}\: (13,0)$
   \begin{itemize}
      \item $\tau = -\frac{3}{8}$, $p = \frac{5}{16} = 0,3125$
      \item $c_0 + c_1 + c_3 \stackrel{p}{\approx} a_0 + a_1 + a_2 + a_3 + k_2 + k_{11}$
   \end{itemize}
\end{itemize}
that do not completely exhaust the potential of the single S-boxes but
on the other hand activate only one S-box of the second round, and thereby
show the larger potential $\lambda = \frac{9}{64}$. Thus we get a 95\%
success probability with only
\[
     N = \frac{3}{\lambda} = \frac{64}{3} \approx 21
\]
known\index{known plaintext}\index{plaintext!known} plaintexts for determining
one bit.

@@

sage: testtext = "Come at nine"
sage: bintext = txt2bbl(testtext)
sage: binstr = bbl2str(bintext); binstr
'010000110110111101101101011001010010000001100001
011101000010000001101110011010010110111001100101'
sage: testkey = [1,0,0,1,0,1,1,0]
sage: keystr = bbl2str(testkey); keystr
'10010110'
sage: cipher = xor(bintext,testkey)
sage: ciphstr = bbl2str(cipher); ciphstr
'110101011111100111111011111100111011011011110111
111000101011011011111000111111111111100011110011'
\end{verbatim}
\caption{XOR encryption\index{XOR}\index{encryption!XOR}\index{cipher!XOR} in Python/SageMath\index{Python}}\label{Sage-code-bool-xor}
\end{sagecode}

@@

def fsr(f,x,n):
  """Generate a feedback shift register sequence.
  Parameters: Boolean function f, start vector x,
  number n of output bits."""
  u = x
  outlist = []
  for i in range (0,n):
    b = f.valueAt(u)
    c = u.pop()
    u.insert(0,b)
    outlist.append(c)
  return outlist
\end{verbatim}
\caption{A feedback shift register (FSR) in Python/SageMath\index{Python}}\label{Sage-code-bool-fsr}
\end{sagecode}

@@

sage: bits = "1000010111001001"
sage: x = str2bbl(bits)
sage: f = BoolF(x)
sage: start = [0,1,1,1]
sage: bitlist = fsr(f, start, 32)
sage: print(bbl2str(bitlist))
11101010101010101010101010101010
\end{verbatim}
\caption{A (very poor) pseudo-random sequence\index{pseudo-random sequence}
   in Python/SageMath\index{Python}}\label{Sage-code-bool-fsr1}
\end{sagecode}

@@

def lfsr(s,x,n):
  """Generate a linear feedback shift register sequence.
  Parameters: Coefficient vector s, start vector x, number n of
  output bits."""
  l = len(s)
  assert l == len(x), "lfsr_Error: Bad length of start vector."
  u = x                           # in Python use u = x.copy()
  outlist = []
  for i in range (0,n):
    b = binScPr(s, u)
    c = u.pop()
    u.insert(0,b)
    outlist.append(c)
  return outlist
\end{verbatim}
\caption{Defining an LFSR\index{LFSR} in Python/SageMath\index{Python}}\label{Sage-code-bool-lfsr}
\end{sagecode}

@@

sage: coeff = [0,1,1,0,1,0,0,0,0,0,0,0,0,0,0,1]
sage: start = [0,1,1,0,1,0,1,1,0,0,0,1,0,0,1,1]
sage: bitlist = lfsr(coeff, start, 1024)
sage: print(bitlist)
### Visualization
sage: m = matrix(GF(2),32,bitlist)
sage: row = [1]*32
sage: n = matrix(GF(2),32,row*32)  # All entries 1
sage: p = m+n                  # Toggle bits of m ---> 1 = black
sage: p.subdivide(range(0,33),range(0,33))
sage: matrix_plot(p, subdivisions=True)
\end{verbatim}
\caption{A pseudo-random bit sequence\index{pseudo-random sequence}
   in Python/SageMath\index{Python}}\label{Sage-code-bool-lfsr1}
\end{sagecode}

@@

\begin{description}
   \item[Class {\tt LSFR}:] ~
      \begin{description}
         \item[Attributes:] ~
            \begin{itemize}
               \item {\tt length}: the length of the register
               \item {\tt taplist} (constant): the list of coefficients (or taps)
                  that define the bits for feedback
               \item {\tt state} (variable): the state of the register
            \end{itemize}
         \item[Methods:] ~
            \begin{itemize}
               \item {\tt setLength}: define the length (used only implicitly
                  for initialization)
               \item {\tt setTaps}: define the list of taps (used only implicitly
                  for initialization)
               \item {\tt setState}: set the state of the register
               \item {\tt getLength}: output the length
               \item {\tt nextBits}: generate a given number of output bits,
                  and set the next state
            \end{itemize}
      \end{description}
\end{description}
For observing the register a method (generically called {\tt \_\_str\_\_}
in Python\index{Python}) is convenient that outputs the attributes in human-readable form.

@@

But implementing a useful bitstream cipher takes more than just a good random generator:
\begin{itemize}
   \item Message integrity\index{integrity} requires additional means
      such as a combination with a cryptographic hash function\index{hash function}. 
   \item The operational conditions must prevent the reuse of (parts of) the
      key stream\index{key stream} in a reliable way. This means that the
      key management\index{key management} requires utmost prudence.
      A possible approach\index{message key}\footnote{%
         This was a usual approach with the cipher machines of World War II.
      }
      is using a longtime general key that consists of certain inner parameters
      of the random generator, and use the remaining parameters including the
      initial state as one-time message key\index{OTP}.
\end{itemize}

@@

SageMath has several functions for bitblocks\index{bitblock} and Boolean
functions\index{Boolean function}\index{function!Boolean}. These are
scattered over different modules, use different data types, and have
some limitations. We could live with them. Nevertheless in this text
we mostly use our own independent (and free) implementation\footnote{%
   This is known as ``reinventing the wheel''. Forgive it in view of
   intended uniformity and consistency. Working through the functions in the
   SageMath samples should facilitate learning the algorithms. A further aspect:
   Our modules are written in pure Python\index{Python} and may be used without installing
   SageMath.
}.

@@

The SageMath classes and functions defined in the following can be found in the
module \verb:bitciphers.sage:. They were developed with Python\index{Python} 3.4.1\footnote{%
   The only relevant difference with Python 2 is the operator {\tt //} for the
   division of integers.
} under OS X, and tested with SageMath 6.2.

### 2017-01-21 script/en/chapters/introduction.tex
CT Book ed 12: Formal corrections: Got rid of 420 warnings "multiply defined ref"; one overfull hbox; Aug16->Jan17 (within this timeframe also update of the primroot section).
CT Book ed 12: Deleted some comments; added Python to Index.

\par \vskip + 35pt
\noindent Bernhard Esslinger
\par \vskip + 10pt
\noindent Heilbronn/Siegen (Germany), August 2016

@@

SageMath is a comprehensive open-source CAS package which can be used to easily
program the mathematical methods explained in this book. A speciality of
this CAS is, that the scripting language is Python (currently version 2.x).
So in a Sage script, you have after an import statement all functions from the
Python language at your disposal.\\
SageMath becomes more and more the standard CAS system at universities.
\par \vskip + 15pt

>>>>

\par \vskip + 35pt
\noindent Bernhard Esslinger
\par \vskip + 10pt
\noindent Heilbronn/Siegen (Germany), August 2016 + January 2017

@@

SageMath is a comprehensive open-source CAS package which can be used to easily
program the mathematical methods explained in this book. A speciality of
this CAS is, that the scripting language is Python (currently version 2.x).
So in a Sage script, you have after an import statement all functions from the
Python\index{Python} language at your disposal.\\
SageMath becomes more and more the standard CAS system at universities.
\par \vskip + 15pt
